---
title: "Parents and College"
author: "Kieran Healy"
date: '2022-04-08'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

First we load the libraries we'll need.


```{r, message=FALSE}
library(tidyverse)
library(socviz)
library(kableExtra)
library(survey) 
library(srvyr) # Let survey work with tidyverse
library(broom)
library(gssr) # https://kjhealy.github.io/gssr

## Tidy up agegroup
convert_agegrp <- function(x){
    x <- gsub("\\(", "", x)
    x <- gsub("\\[", "", x)
    x <- gsub("\\]", "", x)
    x <- gsub(",", "-", x)
    x <- gsub("-89", "+", x)
    regex <- "^(.*$)"
    x <- gsub(regex, "Age \\1", x)
    x
}


```

```{r, message=FALSE, echo=FALSE, results='hide'}
## Fonts for formatting the plots; not required
## Just comment this chunk out if necessary 
library(showtext)
showtext_auto()

library(myriad)
import_myriad_semi()

theme_set(theme_myriad_semi())

```

# GSS Data
Next, the data.

```{r}
## From gssr
data(gss_all)
gss_all
```

## Subset

We select the subset of variables of interest: `age`, `padeg`, `madeg`, and `degree`, together with the survey weights. 

```{r}
cont_vars <- c("year", "id", "ballot", "age")
cat_vars <- c("padeg", "madeg", "degree")
wt_vars <- c("vpsu",
             "vstrat",
             "oversamp",
             "formwt",              # weight to deal with experimental randomization
             "wtssall",             # main weight variable
             "sampcode",            # sampling error code
             "sample")              # sampling frame and method
my_vars <- c(cont_vars, cat_vars, wt_vars)

## Clean up labeled vars as we go, create compwt
gss_df <- gss_all %>%
  filter(year > 1974 & year < 2021) %>% 
  select(all_of(my_vars)) %>% 
  mutate(across(everything(), haven::zap_missing), # Convert labeled missing to regular NA
         across(all_of(wt_vars), as.numeric),
         across(all_of(cat_vars), as_factor), 
         across(all_of(cat_vars), fct_relabel, tolower),
         across(all_of(cat_vars), fct_relabel, tools::toTitleCase),
         compwt = oversamp * formwt * wtssall)
        
```

## Recoding

- Make a "young" variable, ages 18-30 
- Recode the degree measures into HS or lower, Some college, and BA or higher for the respondent (`resp_coll`), father's education (`pa_coll`), and mother's education (`ma_coll`). 
- Create a measure `pama` flagging if _neither_ parent has a BA or higher. 

```{r}

## Age quartiles
qrts <- quantile(as.numeric(gss_df$age), na.rm = TRUE)

## Age quintiles
quintiles <- quantile(as.numeric(gss_df$age), 
                      probs = seq(0, 1, 0.2), na.rm = TRUE)

gss_df <- gss_df %>% 
  mutate(
    young = case_when(age <= 30 ~ "Aged 18-30", 
                      age > 30 ~ "Aged 31+", 
                      TRUE ~ NA_character_),
    ageq = cut(x = age, breaks = unique(qrts), 
               include.lowest=TRUE),
    ageq =  fct_relabel(ageq, convert_agegrp),
    agequint = cut(x = age, breaks = unique(quintiles), 
                   include.lowest = TRUE),
    agequint = fct_relabel(agequint, convert_agegrp),
    self_coll = fct_collapse(degree,
      `HS or less` = c("less than High School", 
                       "High School"), 
      `Some College` = "Associate/Junior College",
      `College or more` = c("Bachelor's", "Graduate")),
    pa_coll = fct_collapse(padeg,
      `HS or less` = c("less than High School", 
                       "High School"), 
      `Some College` = "Associate/Junior College",
      `College or more` = c("Bachelor's", "Graduate")),
    ma_coll = fct_collapse(madeg,
      `HS or less` = c("less than High School", 
                       "High School"), 
      `Some College` = "Associate/Junior College",
      `College or more` = c("Bachelor's", "Graduate")),
    pama = case_when(
      pa_coll != "College or more" & ma_coll != "College or more" ~ "First Gen", 
      pa_coll == "College or more" | ma_coll == "College or more" ~ "College Parent",
      TRUE ~ NA_character_
    )
  )

```

## Unweighted counts

```{r}
## Let's just pick one year

# father
gss_df %>% 
  filter(year == 2010) %>% 
  count(padeg) %>% 
  kbl() %>%
  kable_styling(full_width = F)
```

```{r}
# father recode
gss_df %>% 
  filter(year == 2010) %>% 
  count(pa_coll) %>% 
  kbl() %>%
  kable_styling(full_width = F)

```


```{r}
# mother
gss_df %>% 
  filter(year == 2010) %>% 
  count(madeg) %>% 
  kbl() %>%
  kable_styling(full_width = F)

```

```{r}
# mother recode
gss_df %>% 
  filter(year == 2010) %>% 
  count(ma_coll) %>% 
  kbl() %>%
  kable_styling(full_width = F)

```

```{r}
# respondent
gss_df %>% 
  filter(year == 2010) %>% 
  count(degree) %>% 
  kbl() %>%
  kable_styling(full_width = F)

```

```{r}
# respondent recode
gss_df %>% 
  filter(year == 2010) %>% 
  count(self_coll) %>% 
  kbl() %>%
  kable_styling(full_width = F)
```

## `pama` variable

Check 1:

```{r}
## Neither parent has a college degree
gss_df %>% 
  filter(year == 2010) %>% 
  count(pa_coll, ma_coll) %>% 
  filter(pa_coll != "College or more" & ma_coll != "College or more") %>% 
  summarize(n = sum(n))

## Neither parent has a college degree
gss_df %>% 
  filter(year == 2010) %>% 
  count(pama) %>% 
  filter(pama == "First Gen")
```

Check 2: 

```{r}
## At least one parent has a college degree
gss_df %>% 
  filter(year == 2010) %>% 
  count(pa_coll, ma_coll) %>% 
  filter(pa_coll == "College or more" | ma_coll == "College or more") %>% 
  summarize(n = sum(n))

## At least one parent has a college degree
gss_df %>% 
  filter(year == 2010) %>% 
  count(pama) %>% 
  filter(pama == "College Parent")

```

## Survey-weighted estimates

```{r}
options(survey.lonely.psu = "adjust")
options(na.action="na.pass")

## Before 1975 vpsus are not available
gss_svy <- gss_df %>%
  filter(year > 1974) %>%  
  mutate(stratvar = interaction(year, vstrat)) %>%
  as_survey_design(id = vpsu,
                     strata = stratvar,
                     weights = compwt,
                     nest = TRUE)
gss_svy
```

## Respondents aged 18-30

```{r, fig.width = 7, fig.height = 4}

out <- gss_svy %>%
  filter(young == "Aged 18-30") %>% 
  drop_na(pama) %>%  
  group_by(year, pama) %>%
  summarize(prop = survey_mean(na.rm = TRUE, 
                               vartype = "ci")) %>% 
  filter(pama == "First Gen", year < 2020) %>% 
  ggplot(mapping = aes(x = year, 
                       y = prop, 
                       ymin = prop_low, 
                       ymax = prop_upp)) + 
  geom_ribbon(color = NA, 
              fill = "#0072B2", 
              alpha = 0.4) + 
  geom_line(color = "#0072B2", 
            size = 1.2) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), 
                     limits = c(0, 1)) + 
  labs(title = "Percent of respondents aged 18-30 where neither parent had a college degree ",
       subtitle = "Amongst respondents providing information about both parents.\nPopulation-weighted estimates and confidence intervals.", 
       caption = "Data: GSS. Figure: @kjhealy", 
       x = "Year", y = "Percent")

print(out)

ggsave(filename = "figures/young_pama_nodegree.pdf", out, width = 7, height = 4)
```

## Respondents with Bachelor's Degrees

```{r, fig.width = 7, fig.height = 4.5}
out <- gss_svy %>%
  drop_na(pama) %>% 
  filter(self_coll == "College or more") %>% 
  group_by(year, self_coll, pama) %>%
  summarize(prop = survey_mean(na.rm = TRUE, 
                               vartype = "ci")) %>% 
  filter(pama == "First Gen", 
         year < 2020) %>% 
  ggplot(mapping = aes(x = year, 
                       y = prop, 
                       ymin = prop_low, 
                       ymax = prop_upp)) + 
  geom_ribbon(color = NA, 
              fill = "#0072B2", 
              alpha = 0.4) + 
  geom_line(color = "#0072B2", 
            size = 1.2) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), 
                     limits = c(0, 1)) + 
  labs(title = "Percent of GSS respondents who have a Bachelor's degree,\nand neither of whose parents had a Bachelor's degree",
       subtitle = "Amongst respondents providing information about both parents.\nPopulation-weighted estimates and confidence intervals.", 
       caption = "Data: General Social Survey. Figure: @kjhealy", 
       x = "Year", y = "Percent")
print(out)

ggsave(filename = "figures/firstgen_ba.pdf", out, width = 7, height = 4.5)

```


## By age quintiles 

```{r}
out <- gss_svy %>%
  drop_na(pama) %>%  
  group_by(year, agequint, pama) %>%
  summarize(prop = survey_mean(na.rm = TRUE, 
                               vartype = "ci")) %>% 
  filter(pama == "First Gen", year < 2020) %>% 
  drop_na() %>% 
  ggplot(mapping = aes(x = year, 
                       y = prop, 
                       ymin = prop_low, 
                       ymax = prop_upp)) + 
    geom_hline(yintercept = 0.75, linetype = "dotted", 
             color = "gray40") + 
  geom_hline(yintercept = 0.5, linetype = "dotted", 
             color = "gray40") + 
    geom_hline(yintercept = 0.25, linetype = "dotted", 
             color = "gray40") + 
  geom_ribbon(color = NA, 
              fill = "#0072B2", 
              alpha = 0.4) + 
  geom_line(color = "#0072B2", 
            size = 1.2) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), 
                     limits = c(0, 1)) + 
  facet_wrap(~ agequint, nrow = 1) + 
  labs(title = "Percent of respondents where neither parent had a college degree ",
       subtitle = "Amongst respondents providing information about both parents. Population-weighted estimates and confidence intervals.", 
       caption = "Data: GSS. Figure: @kjhealy", 
       x = "Year", y = "Percent")

print(out)

ggsave(filename = "figures/firstgen_agequnts.pdf", out, width = 12, height = 5)


```

# By age quintiles with Bachelor's degrees

```{r}
## By age quintiles 

out <- gss_svy %>% 
  filter(self_coll == "College or more") %>% 
  drop_na(pama) %>% 
  group_by(year, agequint, self_coll, pama) %>%
  summarize(prop = survey_mean(na.rm = TRUE, 
                               vartype = "ci")) %>% 
  filter(year > 1979, year < 2020) %>% 
  filter(self_coll == "College or more",
         pama == "First Gen") %>% 
  drop_na() %>% 
  ggplot(mapping = aes(x = year, 
                       y = prop, 
                       ymin = prop_low, 
                       ymax = prop_upp)) + 
    geom_hline(yintercept = 0.75, linetype = "dotted", 
             color = "gray40") + 
  geom_hline(yintercept = 0.5, linetype = "dotted", 
             color = "gray40") + 
    geom_hline(yintercept = 0.25, linetype = "dotted", 
             color = "gray40") + 
  geom_ribbon(color = NA, 
              fill = "#0072B2", 
              alpha = 0.4) + 
  geom_line(color = "#0072B2", 
            size = 1.2) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), 
                     limits = c(0, 1)) + 
  facet_wrap(~ agequint, nrow = 1) + 
  labs(title = "Percent of college graduates where neither parent had a college degree, by age quintile",
       subtitle = "Amongst respondents providing information about both parents. Population-weighted estimates and confidence intervals.", 
       caption = "Data: GSS. Figure: @kjhealy", 
       x = "Year", y = "Percent")

print(out)

ggsave(filename = "figures/firstgen_agequnts.pdf", out, width = 12, height = 5)


```


